# The Chi-Square Test {#sec-chisquare}

In previous [sections](https://mathieuferry.github.io/SoMixQuantTutorial/statbis.html), we used cross-tabulations to describe the relationship between two categorical variables. But how do we know whether an observed association is **statistically significant** — i.e., unlikely to be due to chance alone? This is what the **chi-square test** (χ²) tells us.

::: callout-note
## An interactive introduction

Before reading further, we recommend exploring this interactive tool which illustrates the logic of the chi-square test visually: <https://istats.shinyapps.io/ChiSquaredTest/>
:::

------------------------------------------------------------------------

## How Does the Chi-Square Test Work?

The chi-square test compares what we **observe** in the data to what we would **expect** if there were no association between the two variables (the null hypothesis of independence).

For each cell of the cross-tabulation, the expected count is:

$$E_{ij} = \frac{\text{Row total}_i \times \text{Column total}_j}{\text{Grand total}}$$

The test statistic is then:

$$\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}$$

where $O_{ij}$ is the observed count and $E_{ij}$ the expected count in each cell. The larger the discrepancy between observed and expected, the larger the χ² statistic, and the less likely the association is due to chance.

The p-value is derived from comparing this statistic to a chi-square distribution with $(r-1)(c-1)$ degrees of freedom (d.f.), where $r$ is the number of rows and $c$ the number of columns.

::: callout-warning
## The chi-square test is sensitive to sample size

A very large sample will almost always produce a significant chi-square, even for trivially small associations. Conversely, a very small sample may fail to detect a real association. This is why the chi-square test should be calculated on *normalized* survey weights. It can be accompanied by a measure of **effect size** (such as Cramér's V) and a careful look at the **residuals** to understand *where* the association lies.
:::

------------------------------------------------------------------------

## Setup

We continue with the HIES data, using our binary variable `walk` (whether a child walks to school) and examining its association with household wealth, parental education, province, and sex.

```{r}
#| include: false
Sys.setlocale("LC_MESSAGES", "en_US.UTF-8")
```

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(questionr)
library(kableExtra)
library(survey)
library(srvyr)
library(gtsummary)

setwd("/home/groups/3genquanti/SoMix/HIES for workshop")
edu <- readRDS("edu.rds")

edu_school <- edu |>
  mutate(walk = if_else(transport_medium == 1, 1, 0)) |>
  filter(!is.na(distance))
```

------------------------------------------------------------------------

## Method 1: The Simple Approach

This approach follows the logic introduced in the [cross-tabulation section](https://mathieuferry.github.io/SoMixQuantTutorial/statbis.html#a-practical-application-of-the-cross-tabulation) of the tutorial.

### Step 1: Row percentages

```{r}
#| message: false
#| warning: false

edu_school |> 
  freqtable(hhwealthcat, walk, weights = finalweight_25per) |>
  rprop() |>
  kbl(digits = 0) |> 
  kable_classic(full_width = FALSE)
```

### Step 2: Normalize the weights before running the test

The chi-square test is based on weighted counts. If we feed it the raw survey weights, the sample size becomes artificially inflated, which will produce a significant result almost by construction.

The solution is to **normalize** the weights so that they sum to the actual sample size:

```{r}
#| message: false
#| warning: false

edu_school <- edu_school |>
  mutate(weightnorm = finalweight_25per / sum(finalweight_25per) * n())

# Check: normalized weights should sum to the sample size
sum(edu_school$weightnorm)
```

### Step 3: Run the chi-square test

```{r}
#| message: false
#| warning: false

tabfreqwn <- edu_school |> 
  freqtable(hhwealthcat, walk, weights = weightnorm)

chi <- chisq.test(tabfreqwn)
chi
```

### Step 4: Inspect observed, expected, and residuals

Beyond the p-value, the chi-square test produces useful diagnostics that tell us *where* the association is strongest:

```{r}
# Observed counts (weighted)
chi$observed

# Expected counts under independence
chi$expected

# Pearson residuals: (observed - expected) / sqrt(expected)
chi$residuals

# Standardized residuals: values > |2| indicate cells that deviate significantly
chi$stdres
```

::: callout-note
## Reading the standardized residuals

Standardized residuals greater than **+2** indicate cells where the observed count is significantly *higher* than expected; values below **-2** indicate cells where it is significantly *lower*. This tells us exactly which combinations of categories drive the overall association.

For example, if `Richest × walk=0` has a standardized residual of +13.3, it means that children from the richest households are far more likely *not* to walk to school than we would expect under independence.
:::

------------------------------------------------------------------------

## Method 2: Using the Survey Design

The approach above uses normalized weights, which is a pragmatic solution. The more rigorous approach is to use `svychisq()` from the `survey` package, which correctly accounts for the complex sampling design. Here, we do not need to use normalized sampling weights, the function will do it for us.

### Define the survey design

```{r}
#| message: false
#| warning: false

design <- edu_school |> 
  as_survey_design(weights = finalweight_25per)
```

### Descriptive table with `tbl_svysummary`

```{r}
#| message: false
#| warning: false

design |> 
  mutate(walk = factor(walk)) |>
  tbl_svysummary(
    include    = hhwealthcat,
    by         = walk,
    statistic  = all_categorical() ~ "{p}% ({n_unweighted} obs.)",
    percent    = "row"
  ) |> 
  add_overall(last = TRUE) |> 
  modify_header(
    all_stat_cols() ~ "**{level}** ({n_unweighted} obs.)"
  )
```

### Survey-corrected chi-square test

```{r}
#| message: false
#| warning: false

svychisq(~hhwealthcat + walk, 
         design    = design, 
         statistic = "Chisq")
```

::: callout-note
## `chisq.test` vs `svychisq`

`svychisq()` uses a **Rao-Scott correction** that adjusts for the survey design (weights, clustering, stratification).
:::

### Combine table and test in one step

`gtsummary` allows us to add the p-value from `svychisq` directly to the summary table with `add_p()`:

```{r}
#| message: false
#| warning: false

design |> 
  mutate(walk = factor(walk)) |>
  tbl_svysummary(
    include    = hhwealthcat,
    by         = walk,
    statistic  = all_categorical() ~ "{p}% ({n_unweighted} obs.)",
    percent    = "row",
    missing_stat = "{N_miss_unweighted} obs."
  ) |> 
  add_overall(last = TRUE) |> 
  add_p() |>
  modify_header(
    all_stat_cols() ~ "**{level}** ({n_unweighted} obs.)"
  )
```

------------------------------------------------------------------------

## Testing Several Variables Simultaneously

One of the great advantages of `tbl_svysummary` is that we can test the association of `walk` with **multiple variables at once**, each with its own chi-square test:

```{r}
#| message: false
#| warning: false

design |> 
  mutate(walk = factor(walk)) |>
  tbl_svysummary(
    include    = c(hhwealthcat, edu_parent, province, sex),
    by         = walk,
    statistic  = all_categorical() ~ "{p}% ({n_unweighted} obs.)",
    percent    = "row",
    missing_stat = "{N_miss_unweighted} obs."
  ) |> 
  add_overall(last = TRUE) |> 
  add_p() |>
  modify_header(
    all_stat_cols() ~ "**{level}** ({n_unweighted} obs.)"
  )
```

This table gives us a clear overview of which variables are significantly associated with walking to school — and serves as a natural **bivariate screening** before running the multivariate logistic regression of Session 3.

::: callout-caution
## Exercise

1.  Run the chi-square test for the association between `edu_parent` and `walk`. Examine the standardized residuals — which parental education categories show the strongest deviation from independence?

2.  Is the association between `sex` and `walk` statistically significant?

3.  Run `tbl_svysummary` with `add_p()` for the variables `hhwealthcat`, `edu_parent`, `province`, and `sex` on the likelihood of going to school by bus. Comment.

4.  **Bonus**: Compute Cramér's V for the association between `hhwealthcat` and `walk` using the `rcompanion` package (`cramerV(tabfreqwn)`). How does the effect size compare to what the p-value alone suggested?
:::
